Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-15 manuscript/manuscript.tex
Rearrange intro. (Need to re-write to condense/smooth)

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are toy models, which are used to study how historical forces shape genetic diversity, and which form the basis of parametric inference methods.
However, populations are complicated and, moreover, varied in their complications.
No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

@@

In its simplest form, the Kingman coalescent has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
Under a variety of conditions, the coalescent rate is inversely proportional to the size of the population~\autocite{Kingman1982b}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
This observation is the basis for a variety of methods to infer historical population sizes from genetic data.

With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
(Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)

A serious problem for this inference procedure is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, \cite{SchriderEtAl2016} recently demonstrated how population-size inference can be distorted by selective sweeps.
This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce overall genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.

One possible approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but, as with other ``exact'' inference methods, it does not scale to genomic data.)
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

@@

Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby sites---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

\section*{Definitions and previously known results}

\subsection*{The SFS and 2-SFS}
Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.

@@

@@

@@

>>>>

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are simplified mathematical models, which connect historical quantities such as the population size to observable features of genetic data, and thus form the basis of parametric inference methods.
However, populations are complicated and, moreover, varied in their complications.
No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

@@

The Kingman coalescent underlies a variety of methods for inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
Under a variety of conditions, the coalescent rate is inversely proportional to the population size, $N$~\autocite{Kingman1982b}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Observable properties of genetic data depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains some features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or pattern of heterozygocity along the chromosome (e.g., \cite{LiDurbin2011}).

A serious problem for this class of inference procedures is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, when a beneficial mutation increases rapidly in frequency, it distorts the gene genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
When such selective sweeps are common, a large fraction of coalescence events may occur during sweeps.
In this case, the average coalescent rate is proportional to the beneficial mutation supply rate, which is itself \emph{directly} rather than inversely proportional to $N$.
Thus, the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.

While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.

In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.

It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{SelectionScanReview}).
These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic history.
Genomic regions that are outliers from this distribution are presumed to be under selection.
In contrast, we are interested in detecting when the genome-wide background \emph{is not} well-modeled by the Kingman coalescent.

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

@@

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

[STREAMLINE THE FOLLOWING:]

Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.

@@

[DEFINITION FIRST, THEN EXPLANATION]

@@

\section*{Supplementary Methods}
[IF NECESSARY]

@@

% With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
% One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
% The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
% Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
% Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
% Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
% (Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)
